---
title: "Environmental variable download, cleaning, and analysis"
author: "Tanner C. Myers, Pietro L. H. de Mello, Paul M. Hime, and Richard E. Glor"
date: "8/22/2023"
output: html_document
---


# Load R packages
```{r}
work_dir <- "/scratch/tcm0036/distichus-ddRAD/analyses/AnoDist/gea"
setwd(work_dir)

library(sf)
library(terra)
library(sp)
library(geodata)
library(tidyverse)
library(ade4)
library(factoextra)
library(cowplot)
library(grid)
library(gridExtra)

# Load shapefile of Hispaniola for masking rasters
hispaniola <- st_read(dsn = "~/distichus-spatial-variation/data/shape-files/")
```

# Download, crop, and mask CHELSA climate raster layers
```{r}
# Make directories
dir.create("env-data")
dir.create("env-data/raw-vars")
dir.create("env-data/raw-vars/chelsa")
dir.create("env-data/masked")
dir.create("env-data/masked/chelsa")

# Download raw files available here: https://chelsa-climate.org/bioclim/
# NOTE: Be careful, these comprise 48 files and 13 total gigabytes of data
#system("wget -i ~/distichus-ddRAD/gea/envidatS3paths.txt -P /scratch/tcm0036/distichus-ddRAD/analyses/AnoDist/gea/env-data/raw-vars/chelsa/")

# Load downloaded CHELSA raster layers into R
#env <- rast(list.files("env-data/raw-vars/chelsa", pattern = "CHELSA", full.names = TRUE))
## extents differ? Load different variable types and ID culprit
bioclim <- rast(list.files("env-data/raw-vars/chelsa/", pattern = "_bio", full.names = TRUE))
sfc <- rast(list.files(path = "env-data/raw-vars/chelsa/", pattern = "_sfcWind", full.names = TRUE))
hurs <- rast(list.files(path = "env-data/raw-vars/chelsa/", pattern = "_hurs", full.names = TRUE))
clt <- rast(list.files(path = "env-data/raw-vars/chelsa/", pattern = "_clt", full.names = TRUE)) ## Ding, ding, ding!
cmi <- rast(list.files(path = "env-data/raw-vars/chelsa", pattern = "_cmi", full.names = TRUE))
vpd <- rast(list.files(path = "env-data/raw-vars/chelsa", pattern = "_vpd", full.names = TRUE))
pet <- rast(list.files(path = "env-data/raw-vars/chelsa", pattern = "_pet", full.names = TRUE))
rsds <- rast(list.files(path = "env-data/raw-vars/chelsa", pattern = "_rsds", full.names = TRUE)) # same extent, but produces warning
swb <- rast(list.files(path = "env-data/raw-vars/chelsa", pattern = "_swb", full.names = TRUE))

glob_clim <- c(bioclim, pet, rsds, cmi, hurs, swb, sfc, vpd)
clt <- terra::resample(x = clt, y = glob_clim) # fix reduced resolution of clt rasters
glob_clim <- c(glob_clim, clt)

crop_clim <- crop(glob_clim, ext(hispaniola))
mask_clim <- mask(crop_clim, hispaniola)

# write masked rasters to files
lapply(names(mask_clim), function(x) {
    writeRaster(mask_clim[[x]], paste0("env-data/masked/chelsa/", x, ".asc"), overwrite = TRUE)
})
```

# Prepare elevation raster at same resolution (30 arc seconds/1 km)
```{r}
# Download elevation data from `geodata" package
hti_alt <- elevation_30s(country="HTI", path = "env-data/raw-vars/")
dom_alt <- elevation_30s(country = "DOM", path = "env-data/raw-vars/")
hisp_alt <- merge(hti_alt, dom_alt)

hisp_alt <- mask(hisp_alt, hispaniola)

writeRaster(hisp_alt, "env-data/masked/elevation_30sec.asc", overwrite = TRUE)
```

# Prepare mean wet and dry season vegetation index rasters
```{r}
# Create directories
dir.create("env-data/raw-vars/modis")
dir.create("env-data/raw-vars/modis/may")
dir.create("env-data/raw-vars/modis/march")
dir.create("env-data/masked/modis")

# Wet season (should have 11 files)
may_evi <- rast(list.files(path = "env-data/raw-vars/modis/may", pattern = "_monthly_EVI_doy20", full.name = TRUE))
may_ndvi <- rast(list.files(path = "env-data/raw-vars/modis/may", pattern = "_monthly_NDVI_doy20", full.names = TRUE))

# Dry season (should have 11 files)
march_evi <- rast(list.files(path = "env-data/raw-vars/modis/march", pattern = "_monthly_EVI_doy20", full.names = TRUE))
march_ndvi <- rast(list.files(path = "env-data/raw-vars/modis/march", pattern = "_monthly_NDVI_doy20", full.names = TRUE))

# estimate mean for each layer that represent a year
may_evi_mean <- app(may_evi, mean)
may_ndvi_mean <- app(may_ndvi, mean)

march_evi_mean <- app(march_evi, mean)
march_ndvi_mean <- app(march_ndvi, mean)

# now, mask
wet_evi <- mask(may_evi_mean, hispaniola)
wet_ndvi <- mask(may_ndvi_mean, hispaniola)
dry_evi <- mask(march_evi_mean, hispaniola)
dry_ndvi <- mask(march_ndvi_mean, hispaniola)

# For ease of writing to files, concatenate into single SpatRaster
modis <- c(wet_evi, wet_ndvi, dry_evi, dry_ndvi)
names(modis) <- c("wet_season_EVI", "wet_season_NDVI", "dry_season_EVI", "dry_season_NDVI")

lapply(names(modis), function(x){
    writeRaster(modis[[x]], paste0("env-data/masked/modis/", x, ".asc"), overwrite = TRUE)
})
```

# Extract values from rasters
```{r}
 RAD_data <- read_table("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/info/distichus-species-group-Hispaniola-only-popmap-cleaned.tsv", 
                col_names = c("specimen", "specimen_pop", "filepath", "subspecies", "population", "complex", "island", "locality", "longitude", "latitude", "note"))

coords <- RAD_data[, c("longitude", "latitude")]

# Extract values
clim <- as_tibble(cbind(RAD_data$locality, coords, terra::extract(mask_clim, coords, ID = FALSE)))
veg <- as_tibble(cbind(RAD_data$locality, coords, terra::extract(modis, coords, ID = FALSE)))
alt <- as_tibble(cbind(RAD_data$locality, coords, terra::extract(hisp_alt, coords, ID = FALSE)))

# There are NAs -- impute only cells with NAs separately for each source of data
#probs <- distinct(rbind(clim[!complete.cases(clim), 1:3], alt[!complete.cases(alt), 1:3], veg[!complete.cases(veg), 1:3]))
imp_clim <- as_tibble(cbind(clim[!complete.cases(clim), 1:3], terra::extract(mask_clim, clim[!complete.cases(clim), 2:3], ID = FALSE, method = "bilinear")))
imp_alt <- as_tibble(cbind(alt[!complete.cases(alt), 1:3], terra::extract(hisp_alt, alt[!complete.cases(alt), 2:3], ID = FALSE, method = "bilinear")))
imp_veg <- as_tibble(cbind(veg[!complete.cases(veg), 1:3], terra::extract(modis, veg[!complete.cases(veg), 2:3], ID = FALSE, method = "bilinear")))

clim[!complete.cases(clim), 4:51] <- imp_clim[, 4:51]
veg[!complete.cases(veg), 4:7] <- imp_veg[, 4:7]
alt[!complete.cases(alt), 4] <- imp_alt[, 4]

env_df <- as_tibble(data.frame(cbind(clim, veg[, -c(1:3)], alt[, -c(1:3)])))

# Rename columns to simplify
new_names <- colnames(env_df) %>% str_replace("CHELSA_", "") %>% str_replace("_1981.2010_V.2.1", "")
colnames(env_df) <- new_names
write.table(env_df, file = "env-data/environmental-variables-raw.tsv", col.names = TRUE, row.names = FALSE, quote = FALSE)
```

# Assess variables for normality and transform
```{r}
sw <- apply(env_df[4:ncol(env_df)], 2, shapiro.test)

sig <- lapply(names(sw), function(x){
        sw[[x]]$p.value > 0.05
})

# log-transform everything
log_env <- apply(env_df[4:ncol(env_df)], 2, function(x){
    if (min(x) < 0){
        log(x + abs(min(x)) + 1) # Some variables have negative values
    } else {
        log(x)
    }
})

unscaled_env <- cbind(env_df[, 1:3], log_env)
full_env <- as_tibble(cbind(env_df[, 1:3], scale(unscaled_env[, 4:ncol(unscaled_env)])))
write.table(full_env, "env-data/environment-log-scaled.tsv", col.names = TRUE, row.names = FALSE, quote = FALSE)
```

## Figure 6 - SNP correlation variables
```{r}
alt <- rast("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/AnoDist_aln/gea/env-data/masked/elevation_30sec.asc")
bio4 <- rast("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/AnoDist_aln/gea/env-data/masked/CHELSA_bio4_1981-2010_V.2.1.asc")
bio18 <- rast("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/AnoDist_aln/gea/env-data/masked/CHELSA_bio18_1981-2010_V.2.1.asc")
pal_grays <- hcl.colors(1000, "Grays")
pal_grays <- rev(pal_grays)

# altitude
ggplot(data = RAD_data, aes(x = longitude, y = latitude)) + 
        geom_spatraster(data = alt, alpha = 1, show.legend = FALSE) + scale_fill_gradientn(colours = pal_grays, na.value = NA) +
        geom_point(aes(color = cluster), size = 8, show.legend = FALSE) +
        scale_color_manual(values = c("dodgerblue1", "sienna", "yellow", "gray80", "orangered")) +
        geom_point(size = 8, shape = 21, color = "black") + #theme_light() + theme(axis.title = element_blank())
        theme(panel.background = element_rect(fill = "white"), axis.title = element_blank(),
            axis.ticks.x = element_blank(), axis.ticks.y = element_blank(),
            axis.text.x = element_blank(), axis.text.y = element_blank())
ggsave("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/ms-figures/alt.pdf")      

# bio4
ggplot(data = RAD_data, aes(x = longitude, y = latitude)) + 
        geom_spatraster(data = bio4, alpha = 1, show.legend = FALSE) + scale_fill_gradientn(colours = pal_grays, na.value = NA) +
        geom_point(aes(color = cluster), size = 8, show.legend = FALSE) +
        scale_color_manual(values = c("dodgerblue1", "sienna", "yellow", "gray80", "orangered")) +
        geom_point(size = 8, shape = 21, color = "black") + #theme_light() + theme(axis.title = element_blank())
        theme(panel.background = element_rect(fill = "white"), axis.title = element_blank(),
            axis.ticks.x = element_blank(), axis.ticks.y = element_blank(),
            axis.text.x = element_blank(), axis.text.y = element_blank())
ggsave("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/ms-figures/bio4.pdf")

# BIO18
ggplot(data = RAD_data, aes(x = longitude, y = latitude)) + 
        geom_spatraster(data = bio18, alpha = 1, show.legend = FALSE) + scale_fill_gradientn(colours = pal_grays, na.value = NA) +
        geom_point(aes(color = cluster), size = 8, show.legend = FALSE) +
        scale_color_manual(values = c("dodgerblue1", "sienna", "yellow", "gray80", "orangered")) +
        geom_point(size = 8, shape = 21, color = "black") + #theme_light() + theme(axis.title = element_blank())
    theme(panel.background = element_rect(fill = "white"), axis.title = element_blank(),
        axis.ticks.x = element_blank(), axis.ticks.y = element_blank(),
        axis.text.x = element_blank(), axis.text.y = element_blank())
ggsave("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/ms-figures/bio18.pdf")
```

# Run PCA on environment variables
```{r}
data <- as_tibble(cbind(RAD_data, env_df[, 4:ncol(env_df)]))
pca <- dudi.pca(data[, 12:ncol(data)], center = TRUE, scale = TRUE, scannf = FALSE, nf = 3)

round(100*(pca$eig/sum(pca$eig)), 2) # variation explained
pve <- fviz_eig(pca, addlabels = TRUE, ylim = c(0, 45))

    pca$c1 # loadings

# Use population structure figure cluster assignments to color sampling localities
adm_Q <- read_table("~/Dropbox/Distichus_Project/ddRADseq_Phylogeography/AnoDist_aln/population-structure/ADMIXTURE_PCA/geno90/distichus_geno90_filtered_LD_50kb_100_0.1_ready.5.Q", col_names = FALSE)
    cluster <- apply(adm_Q, 1, which.max)

pc_df <- as_tibble(cbind(RAD_data[, c("specimen_pop", "locality", "longitude", "latitude")], cluster, pca$li))

myColors <- c("dodgerblue1", "sienna", "yellow", "gray80", "orangered")
col_labs <- c("South Paleo-island", "Eastern A. d. dominicensis", "A. d. ravitergum", "A. d. properus", "A. d. ignigularis")

p12 <- ggplot(data = pc_df, aes(x = Axis1, y = Axis2, color = as.factor(cluster), alpha = 0.85)) +
    geom_point(cex = 3, show.legend = FALSE) + stat_ellipse(show.legend = FALSE) +
    scale_color_manual(values = myColors, labels = col_labs,
        guide = guide_legend(label.theme = element_blank())) +
    guides(alpha = "none") +
    labs(color = "Cluster", x = "Axis 1 (39.59% of variation)", y = "Axis 2 (17.57% of variation)") + # UDPATE values if needed
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 3),
    axis.title = element_text(size = 18, colour = "black", family = "Helvetica", face = "bold")) +
    theme_bw()
p23 <- ggplot(data = pc_df, aes(x = Axis2, y = Axis3, color = as.factor(cluster), alpha = 0.85)) +
    geom_point(cex = 3, show.legend = FALSE) + stat_ellipse(show.legend = FALSE) +
    scale_color_manual(values = myColors, labels = col_labs,
        guide = guide_legend(label.theme = element_text(angle = 0, face = "italic"))) +
    guides(alpha = "none") +
    labs(color = "Cluster", x = "Axis 2 (17.57% of variation)", y = "Axis 3 (11.59% of variation)") + # UDPATE values if needed
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 3),
    axis.title = element_text(size = 18, colour = "black", family = "Helvetica", face = "bold")) +
    theme_bw()
p13 <- ggplot(data = pc_df, aes(x = Axis1, y = Axis3, color = as.factor(cluster), alpha = 0.85)) +
    geom_point(cex = 3, show.legend = FALSE) + stat_ellipse(show.legend = FALSE) +
    scale_color_manual(values = myColors, labels = col_labs,
        guide = guide_legend(label.theme = element_blank())) +
    guides(alpha = "none") +
    labs(color = "Cluster", x = "Axis 2 (39.59% of variation)", y = "Axis 3 (11.59% of variation)") + # UDPATE values if needed
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 3),
    axis.title = element_text(size = 18, colour = "black", family = "Helvetica", face = "bold")) +
    theme_bw()
row1 <- plot_grid(p12, p13, ncol = 2, nrow = 1)
row2 <- plot_grid(p23, legend, pve, rel_widths = c(3, 1, 2), nrow = 1, ncol = 3)
plot_grid(row1, row2, nrow = 2, byrow = TRUE)
ggsave("~/Dropbox/Apps/Overleaf/Anolis distichus Population Genomics ms/supp-mat/environmental-pca.pdf")
```

# Load environmental variables and explore collinearity
```{r}
full_env <- read_table("env-data/environment-log-scaled.tsv")

# Shows scatter plots, histograms, and Pearson"s correlation coefficients. Really ugly for this number of variables
pdf("env-data/pairs-panel-plot.pdf", width = 30, height = 30)
    pairs.panels(full_env[, 4:ncol(full_env)], scale = TRUE)
dev.off()

pdf("env-data/corrplot.pdf", width = 30, height = 30)
corrplot(cor(full_env[, 4:ncol(full_env)]), type = "upper")
dev.off()

# Estimate Variance Influence Factors (VIFs) for each of the predictors
## `vifcor` function of the `usdm` package estimates VIFs by finding
## the pair with a maximum linear correlation and excludes
## the one with the greater VIF until no variable
## pair with a high correlation coefficient remains
vifcor <- usdm::vifcor(x = as.matrix(full_env[, 4:ncol(full_env)]), th = 0.9, method = "pearson")

## `vifstep` function of the `usdm` package estimates VIFs
## all at once, excluding the one with the highest VIF,
## repeating until no variables
## with VIF higher than th remains
vifstep <- usdm::vifstep(x = as.matrix(full_env[, 4:ncol(full_env)]), th = 10, method = "pearson")


sel_env <- as_tibble(cbind(full_env[, 1:3], full_env[, vifstep@results$Variables]))
write.table(sel_env, "selected-environmental-variables.tsv", quote = FALSE, col.names = TRUE, row.names = FALSE)
```

```{r}
sel_env <- read_table("selected-environmental-variables.tsv")
## Filter variables to include only forward-selected environmental variables
ordi_vars <- c("clt_min", "elevation_30sec", "bio4", "bio18", "bio9", "cmi_range", "bio3")
sel_env <- sel_env[ordi_vars]

pca <- dudi.pca(sel_env, center = TRUE, scale = TRUE, scannf = FALSE, nf = 3)
round(100*(pca$eig/sum(pca$eig)), 2) # variation explained
pve <- fviz_eig(pca, addlabels = TRUE, ylim = c(0, 45))

    pca$c1 # loadings
#                        CS1         CS2         CS3
# clt_min         -0.1663960  0.60402914 -0.17738840
# elevation_30sec -0.5173283 -0.21984513 -0.21436439
# bio4            -0.1897373  0.50858056 -0.46324442
# bio18           -0.4231809 -0.03582881  0.58556290
# bio9             0.5318651  0.23777316  0.04820212
# cmi_range       -0.3386439  0.43179788  0.41186898
# bio3            -0.3033831 -0.28964383 -0.43947976

pc_df <- as_tibble(cbind(RAD_data[, c("specimen_pop", "locality", "longitude", "latitude")], cluster, pca$li))
    write.table(pc_df, "fwd-select-env-pc1-3.tsv", quote = FALSE, col.names = TRUE, row.names = FALSE)

myColors <- c("dodgerblue1", "sienna", "yellow", "gray80", "orangered")
clust_labs <- c(expression("West "*italic("A. distichus")), expression("East "*italic("A. d. dominicensis")),
                expression(italic("A. d. ravitergum")), expression(italic("A. d. properus")), expression(italic("A. d. ignigularis")))

p12 <- ggplot(data = pc_df, aes(x = Axis1, y = Axis2, color = as.factor(cluster), alpha = 0.85)) +
    geom_point(cex = 3, show.legend = FALSE) + stat_ellipse(show.legend = FALSE) +
    scale_color_manual(values = myColors, labels = clust_labs,
        guide = guide_legend(label.theme = element_blank())) +
    guides(alpha = "none") +
    labs(color = "Cluster", x = "Axis 1 (34.44% of variation)", y = "Axis 2 (29.07% of variation)") + # UDPATE values if needed
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 3),
    axis.title = element_text(size = 18, colour = "black", family = "Helvetica", face = "bold")) +
    theme_bw()
p23 <- ggplot(data = pc_df, aes(x = Axis2, y = Axis3, color = as.factor(cluster), alpha = 0.85)) +
    geom_point(cex = 3, show.legend = FALSE) + stat_ellipse(show.legend = FALSE) +
    scale_color_manual(values = myColors, labels = clust_labs,
        guide = guide_legend(label.theme = element_text(angle = 0, face = "italic"))) +
    guides(alpha = "none") +
    labs(color = "Cluster", x = "Axis 2 (29.07% of variation)", y = "Axis 3 (15.58% of variation)") + # UDPATE values if needed
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 3),
    axis.title = element_text(size = 18, colour = "black", family = "Helvetica", face = "bold")) +
    theme_bw()
p13 <- ggplot(data = pc_df, aes(x = Axis1, y = Axis3, color = as.factor(cluster), alpha = 0.85)) +
    geom_point(cex = 3, show.legend = FALSE) + stat_ellipse(show.legend = FALSE) +
    scale_color_manual(values = myColors, labels = clust_labs,
        guide = guide_legend(label.theme = element_blank())) +
    guides(alpha = "none") +
    labs(color = "Cluster", x = "Axis 1 (34.44% of variation)", y = "Axis 3 (15.58% of variation)") + # UDPATE values if needed
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 3),
    axis.title = element_text(size = 18, colour = "black", family = "Helvetica", face = "bold")) +
    theme_bw()
row1 <- plot_grid(p12, p13, ncol = 2, nrow = 1)
row2 <- plot_grid(p23, legend, pve, rel_widths = c(3, 1, 2), nrow = 1, ncol = 3)
plot_grid(row1, row2, nrow = 2, byrow = TRUE)
ggsave("fwd-selected-env-pca-biplots.pdf")
```